\Large\textbf{General Stage Untangling with Applications to Algorithm Derivation}

### 2014-06-19 paper/semantics.tex
Minor edits to the semantics section 3.1.
-- Highlight in red statements that need to be changed.
-- Fix choice of numbers in the examples.

Although there is a strict one-way dependence between the two stages at the type
level, stage-\bbone{} and stage-\bbtwo{} terms are defined mutually recursively:
the $\next$ construct embeds stage-\bbtwo{} expressions into stage \bbone,
whereas $\prev$ embeds stage-\bbone{} expressions into stage \bbtwo.
The $\next$ and $\prev$ constructs are the only expressions that interact with the stage of a term,
and we surround their arguments with braces in \lang\ syntax to clearly indicate stage boundaries within a program.

@@

% As our concern is stage splitting of explicitly staged programs expressed in \lang.
We note that our work does not consider the problem of {\em binding time analysis}---transforming an unstaged program into a staged program by inserting appropriate $\next$, $\prev$, and $\pause$ constructs. We assume that binding time analysis is performed manually by a programmer or via automatic analysis of unstaged code to arrive at a valid \lang\ program used as input for subsequent spitting.

%The problem of how to convert an unstaged program into a staged one by inserting these constructs is called {\em binding time analysis}.
%We do not consider that problem here, and instead assume that it has been solved either automatically or by a programmer.

@@

As a simple example of expressing staged programs in \lang\, consider the following fast exponentiation algorithm below, which 
calculates $b^p$ in $\log p$ time:
\[
	\mathit{fexp}(b,p) = \left \{ \begin{array}{ll} 
		1 &  p = 0 \\ 
		\mathit{fexp}(b,p/2)^2 & p \text{ even} \\ 
		b \cdot \mathit{fexp}(b,p-1) & p \text{ odd} \end{array}
	\right .
\]

% Ignoring for the moment how to define recursive functions in \lang,
\noindent
The algorithm is expressed in \lang\ below:%
\footnote{We render $\fut$ as {\tt \$} in code examples.}

@@

Although the code for {\tt fexp} looks very much like the unstaged mathematical
definition above, it is in fact a staged program: the {\tt if}s and exponent
decomposition are all at stage \bbone, since they occur within $\prev$ blocks.
Essentially, the type system is sufficient to prove the observation that all of
the decomposition of the exponent can be done at the stage \bbone~without
requiring us to disrupt the elegant functional structure of the code.

This example also illustrates how stage \bbone{} and stage \bbtwo{} code can be
arbitrarily nested.  Ordinary evaluation always eliminates outermost redexes
first, but this conflicts with the precept that we evaluate all stage \bbone{}
code before stage \bbtwo{} code, in the case that stage \bbone{} code is inside
stage \bbtwo{} code.

A suitably staged dynamic semantics for \lang{} must therefore evaluate
\emph{all} the stage \bbone{} subexpressions of a term before its stage
\bbtwo{} subexpressions.

Consider how we could achieve this in the following stage \bbtwo{} term:
\begin{lstlisting} 
prev{
let x = (next {4+5}, 7+8) in
next{
	prev{#1 x} * prev{#1 x} * prev{hold (#2 x)}
}
}
\end{lstlisting}
One option is to duplicate the contents of the first $\next$ expression, to produce
\begin{lstlisting}
(4+5) * (4+5) * 15
\end{lstlisting}

@@

We take a different approach in our dynamic semantics, which is to let-bind the
contents of the first $\next$ expression to some variable (here, $x1$) and
duplicate that variable,
\begin{lstlisting} 
let x1 = 4+5 in
x1 * x1 * 15
\end{lstlisting}

@@

How do our semantics achieve this behavior? In the original term,
\begin{lstlisting}
prev {
let x = (next {4+5}, 7+8) in
next{
	prev{#1 x} * prev{#1 x} * prev{hold (#2 x)}
}
}
\end{lstlisting}
we must start by evaluating the binding for $x$.  We can't evaluate inside the
$\next$ statement, but must still substitute some result into the body of the
\texttt{let}. We replace the contents of $\next$ with a variable and make an
explicit substitution, which then floats up to the top of the containing
$\prev$:
\begin{lstlisting} 
prev{
[4+5/x1]
let x = (next{x1}, 15) in
next{
	prev{#1 x} * prev{#1 x} * prev{hold (#2 x)}
}
}
\end{lstlisting}
We're now free to perform the stage-\bbone~substitution for $x$.  
\begin{lstlisting} 
prev{
[4+5/x1]
next{
	prev{#1 (next{x1}, 15)} 
	* prev{#1 (next{x1}, 15)} 
	* prev{hold (#2 (next{x1}, 15))}
}
}
\end{lstlisting}
To evaluate the remaining $\next$, we must first partially evaluate the body by finding all of the contained stage-\bbone~terms and reducing them:
\begin{lstlisting} 
prev{
[4+5/x1]
	next{ x1 * x1 * 15 }
}
\end{lstlisting}
The final step in evaluating the outer $\prev$ is reifying the contained substitutions into let statements, yielding
\begin{lstlisting} 
let x1 = 4+5 in
 x1 * x1 * 15
\end{lstlisting}

Once again, the point here is that we did not duplicate the $4+5$ in our residual.  
Instead we lifted it out into a substitution, leaving a hatted variable [\TODO add hats above] in its place,
and we reified the substitution into a let-binding at the enclosing $\prev$.

@@

The algorithm described above creates three different kinds of expressions which
cannot be evaluated further at a particular stage:
\begin{itemize}
\item Partial values ($\pval$s) are stage \bbone{} terms whose stage \bbone{}
subexpressions have all been fully evaluated. In the example above,
\verb|next {x1 * x1 * 15}|.
\item Residuals ($\res$es) are stage \bbtwo{} terms whose stage \bbone{}
subexpressions have all been fully evaluated. In the example above,
\verb|4+5|.
\item Values ($\val$s) are stage \bbtwo{} terms which are fully evaluated; these
are the results of a computation after both stages have been completed. In the
example above, the term evaluates to \verb|1215|.
\end{itemize}
The $\redonesym$ judgment takes an open stage \bbone{} term to a pair of an
environment $\xi$ mapping newly-created variables to stage \bbtwo{} residuals,
and a partial value $v$ which may refer to those variables. This judgment
necessarily traverse stage \bbtwo{} subexpressions in order to find their stage
\bbone{} subexpressions; we call this process \emph{speculation}. Speculation is
implemented by the $\redtwosym$ judgment, which takes a stage \bbtwo{} term to a
residual.

>>>>

\needsfix{Although there is a strict one-way dependence between the two stages at the type
level}, stage-\bbone{} and stage-\bbtwo{} terms are defined mutually recursively:
the $\next$ construct embeds stage-\bbtwo{} expressions into stage \bbone,
whereas $\prev$ embeds stage-\bbone{} expressions into stage \bbtwo.
The $\next$ and $\prev$ constructs are the only expressions that interact with the stage of a term,
and we surround their arguments with braces in \lang\ syntax to clearly indicate stage boundaries within a program.

@@

We note that our work does not consider the problem of {\em binding time analysis}---transforming an unstaged program into a staged \lang\ program by inserting appropriate $\next$, $\prev$, and $\pause$ constructs. We assume that binding time analysis is performed manually by a programmer or via automatic analysis of unstaged code to arrive at a valid \lang\ program used as input for subsequent splitting.

@@

As a simple example of expressing staged programs in \lang, consider the following fast exponentiation algorithm, which 
calculates $b^p$ in $\log p$ time:
\[
	\mathit{fexp}(b,p) = \left \{ \begin{array}{ll} 
		1 &  p = 0 \\ 
		\mathit{fexp}(b,p/2)^2 & p \text{ even} \\ 
		b \cdot \mathit{fexp}(b,p-1) & p \text{ odd} \end{array}
	\right .
\]

% Ignoring for the moment how to define recursive functions in \lang,
\noindent
The stage-\bbone\ term below defines the function {\tt fexp} that implements the fast exponentiation algorithm:%
\footnote{We render $\fut$ as {\tt \$} in code examples.}

@@

Although the code for {\tt fexp} looks very much like the unstaged mathematical
definition above, it is in fact a staged program: the {\tt if} predicates and exponent
decomposition are all stage-\bbone\ terms, since they occur within $\prev$ blocks.
\needsfix{Essentially, the type system is sufficient to prove the observation that all of
the decomposition of the exponent can be done at the stage \bbone\ without
requiring disruption of the elegant functional structure of the code.}

A key attribute of the {\tt fexp} code example is the nesting of stage-\bbone\ and stage-\bbtwo\ expressions. Ordinary term evaluation always eliminates outermost redexes first, 
however in the case that stage \bbone\ expressions are contained inside
stage \bbtwo\ ones (such as in the example above), this strategy conflicts with the precept of staged execution: that all stage \bbone\ code be evaluated before the evaluation of stage \bbtwo\ code. Therefore, a suitably staged dynamic semantics for \lang\ must evaluate
\emph{all} the stage \bbone\ subexpressions of a term before its stage
\bbtwo\ subexpressions.

To understand how this behavior can be achieved, consider the following stage-\bbone\ term:
\begin{lstlisting} 
let x = (next {1+2}, 3+4) in
  next{ prev{#1 x} * prev{#1 x} * prev{hold(#2 x)} }
\end{lstlisting}
One option is to duplicate the contents of the first $\next$ expression, to produce
\begin{lstlisting}
(1+2) * (1+2) * 7
\end{lstlisting}

@@

We take a different approach in our dynamic semantics, which is to let-bind the
contents of the first $\next$ expression to some variable (here, $x1$) and
duplicate that variable,
\begin{lstlisting} 
let x1 = 1+2 in
  x1 * x1 * 7
\end{lstlisting}

@@

How do our semantics achieve this behavior? In the original term,
\begin{lstlisting}
let x = (next {1+2}, 3+4) in
  next{ prev{#1 x} * prev{#1 x} * prev{hold (#2 x)} }
\end{lstlisting}
we must start by evaluating the binding for $x$.  We can't evaluate inside the
$\next$ statement, but must still substitute some result into the body of the
\texttt{let}. We replace the contents of $\next$ with a variable and make an
explicit substitution, which then floats up to the top of the containing
$\prev$:
\begin{lstlisting} 
[1+2/x1]
let x = (next{x1}, 7) in
next{ prev{#1 x} * prev{#1 x} * prev{hold (#2 x)} }
\end{lstlisting}
We're now free to perform the stage-\bbone~substitution for $x$.  
\begin{lstlisting} 
[1+2/x1]
next{
	prev{#1 (next{x1}, 7)} *
	prev{#1 (next{x1}, 7)} *
	prev{hold (#2 (next{x1}, 7))}
}
\end{lstlisting}
To evaluate the remaining $\next$, we must first partially evaluate the body by finding all of the contained stage-\bbone~terms and reducing them:
\begin{lstlisting} 
[1+2/x1]
	next{ x1 * x1 * 7 }
\end{lstlisting}
The final step in evaluating the outer $\prev$ is reifying the contained substitutions into let statements, yielding
\begin{lstlisting} 
let x1 = 1+2 in
  x1 * x1 * 7
\end{lstlisting}

Once again, the point here is that we did not duplicate the $1+2$ in our residual.  
Instead we lifted it out into a substitution, leaving a hatted variable [\TODO add hats above] in its place,
and we reified the substitution into a let-binding at the enclosing $\prev$.

@@

The algorithm described above creates three different kinds of expressions which
cannot be evaluated further at a particular stage:
\begin{itemize}
\item Partial values ($\pval$s) are stage \bbone{} terms whose stage \bbone{}
subexpressions have all been fully evaluated. In the example above,
\verb|next {x1 * x1 * 7}|.
\item Residuals ($\res$es) are stage \bbtwo{} terms whose stage \bbone{}
subexpressions have all been fully evaluated. In the example above,
\verb|1+2|.
\item Values ($\val$s) are stage \bbtwo{} terms which are fully evaluated; these
are the results of a computation after both stages have been completed. In the
example above, the term evaluates to \verb|1215|.
\end{itemize}
The $\redonesym$ judgment takes an open stage \bbone{} term to a pair of an
environment $\xi$ mapping newly-created variables to stage \bbtwo{} residuals,
and a partial value $v$ which may refer to those variables. This judgment
necessarily traverse stage \bbtwo{} subexpressions in order to find their stage
\bbone{} subexpressions; we call this process \emph{speculation}. Speculation is
implemented by the $\redtwosym$ judgment, which takes a stage \bbtwo{} term to a
residual.
